The Rheumatoid Arthritis data  \nSupplementary File for  \n`Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests`

### 2025-05-03 Supplementary_Files/Supplementary_File_Rheumatoid_Data.Rmd
added Supplementary_Files

---
title: "The Rheumatoid Arthritis data  \nSupplementary File for "Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests"
author: "Hawlader Al-Mamun,  Rob Dunne"
date: "Wednesday, September 25, 2024"
output: 
    bookdown::pdf_document2:
     toc: true
     toc_depth: 3
     fig_caption: yes
     citation_package: biblatex
bibliography: 
header-includes:
  \usepackage{dcolumn}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage{biblatex}
  \usepackage{caption}
  \usepackage{subcaption}
---

@@

```
\clearpage
# Notes

- is the data available? What about the preprocessing? Can we include this in a vignette?
- dim(dt)   4798 94485 but the paper says 94,486
- looked at variable importance. Variables that have more splits as they are important are likely to have more consecutive splits. Should we take this into account?
- I removed splits from AllInteractions where the split is on the same variable again. This does not matter for the regression method but might for other methods (Empirical Bayes)
- There is a hotspot in chromosome 6. Did we know about this? Does it get dropped when you filter on LD
- we are looking at the interaction as the product of the two SNP variables, so it takes the values $\{0,1,2,4\}$. so we are threating it as a
  regression but it does not take the value 3.  If we treat them as factors we get levels(d1:d2) "0:0" "0:1" "0:2" "1:0" "1:1" "1:2" "2:0" "2:1" "2:2"
  9 levels. Which is a more reasonable model? 
- I look at using the **empiricalzip** (Empirical Bayes zero inflated Poisson) package to model the interaction counts and set a significance
level. The agreement with the regression testing is not great.
- we have the file "Cleaned_CaseControl_RAauto_pruned200kb_.3.bim"

# load data

@@

```{r, echo=FALSE, eval=FALSE}
library(lmtest)
library(ranger)
#library(RCy3)

#Noise <- 0.1
numIte  <-  1
t <- numIte
ntree <- 1000

@@

dt$pheno <- as.factor(dt$pheno)
#dt <- dt[which(dt$treatment=="short"),-2]

# quick load of results etc
dirOut  <-  paste0("Results")
load(file="Results/var_importance_gini.Rdata")
load(file=paste0(dirOut,"/All_Interactions.Rdata"))
load(file="./Results/interaction_snps_counts_fixed.Rdata")
load(file=paste0(dirOut,"/lm_results_fixed.Rdata"))

@@

```{r, echo=FALSE, eval=FALSE}
##Number of trees:                  1000 
#set.seed(sample(1:1000,1))
set.seed(100)
#system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#89141.264     3.601  1568.711 
#save(model.rf, file="model.rf.Rdata")

##Number of trees:                  5000 
#system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = 5000, mtry = ncol(dt)/3))
# user     system    elapsed 
#443201.110      8.132   7185.540 
#Number of trees:                  5000 
#Sample size:                      4798 
#Number of independent variables:  94484 
#Mtry:                             31495 
#Target node size:                 1 
#Variable importance mode:         none 
#Splitrule:                        gini 
#OOB prediction error:             33.18 % 
#save(model.rf, file="model.rf_temp2.Rdata")

##Number of trees:   1000, importance = 'impurity'
set.seed(100)
#petrichor-i1: salloc -A OD-217714 --mem=100GB --nodes=1 --ntasks-per-node=64    -J interactive -t 6:00:00 srun --pty /bin/bash -l
#etrichor-i1: sacct -j 60755241  --format="jobid,ntasks,Elapsed,TotalCPU,MaxRSS,AveVMSize,MaxVMSize"
#JobID          NTasks    Elapsed   TotalCPU     MaxRSS  AveVMSize  MaxVMSize 
#60755241                00:13:56  00:06.682                                  
#60755241.ex+        1   00:13:56  00:00.001       995K    124080K    124080K 
#60755241.0         64   00:13:53  00:06.680 104545713K 3274946112 192422568K

system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#89397.171     4.807  1591.941 
#save(model.rf, file="model.rf_temp3.Rdata")
var_importance <- model.rf$variable.importance
head(rev(sort(var_importance)))
t2 <-count_variables(model.rf)
names(t2) <-names(var_importance)

@@

```{r RFlocalfdr, eval=TRUE, echo=FALSE, fig.cap='RFlocalfdr',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./script3/RFlocalfdr.png")

@@

```{R, echo=FALSE, eval=FALSE}
library(RFlocalfdr)
vignette(package="RFlocalfdr")
#no vignettes found -- why is this?
imp<-log(var_importance)  #6275 var_importance vales =0 
plot(density(imp),xlab="log importances",main="")

@@

cutoffs <- c(4,5,6)
res.con2 <- determine_cutoff(imp,t2 ,cutoff=cutoffs,plot=c(4,5,6))
plot(c(0,1,2,3,4,5,6),c(res.con[,3],res.con2[,3]),pch=15,col="red",cex=1.5,ylab="max(abs(y - t1))")

@@

png("./script3/RFlocalfdr.png")
aa<-significant.genes(ppp,temp,cutoff=0.05,debug.flag=0,do.plot=TRUE,use_95_q=TRUE)
dev.off()
length(aa$probabilities) # 74

@@

```{r hotspot, eval=TRUE, echo=FALSE, fig.cap='There is a hotspot in chromosome 6',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./figs/hotspot.png")

@@

# count the interactions in actual data

```{R, echo=FALSE,eval=FALSE}
#t2 <- system.time({

  # countMatrix <- matrix(0,ncol = ncol(dt)-1,nrow=ncol(dt)-1)
  # row.names(countMatrix) <- colnames(dt)[-1]
  # colnames(countMatrix) <- colnames(dt)[-1]
  #

  for(i in 1:ntree){
    #tree2 <- getTree(model.rf2,i,T)
    tree <- treeInfo(model.rf,i)
    #tree <- treeInfo(model.n,5)

@@

#saveRDS(All_Interactions,paste0(dirOut,"/All_Interactions.rds"))
#All_Interactions <- readRDS(paste0(dirOut,"/All_Interactions.rds"))
#save(All_Interactions,file=paste0(dirOut,"/All_Interactions.Rdata"))
#dim(All_Interactions) #[1] 294310      2

```
```{R, echo=FALSE, eval=FALSE}
# just some checks on how  All_Interactions is organized. 
head(All_Interactions)
#            X1           X2      294310      2
#1  X6_32471505  X6_32771829
#2 X21_44075267  X6_32471505
#3 X14_84754343  X4_67191986

@@

b

@@

#saveRDS(interaction_snps_counts,paste0(dirOut,"/interaction_snps_counts_fixed.rds"))
#interaction_snps_counts <- readRDS("./Results/interaction_snps_counts_fixed.rds")
#save(interaction_snps_counts, file="./Results/interaction_snps_counts_fixed.Rdata")
#load(file="./Results/interaction_snps_counts_fixed.Rdata")

@@

# data set up

@@

What would the distribution of counts look like if 291142 interactions were assigned randomly?
Under a random assignment, this follows a binomial distribution that can be approximated by a Poisson distribution with:
$\lambda= 291,142 / 4,463,565,886 \approx 0.000065$

-  Count = 1: approximately 290,849 pairs
-  Count = 2: approximately 9.5 pairs
-  Count $\geq 3$: effectively 0 pairs

The observed data shows:

- Count = 1: 290502  pairs
-  Count = 2: 418   pairs
-  Count = 3: 89  pairs
-  Count $\geq 4$: 133 pairs

\clearpage
\FloatBarrier

# empiricalzip

@@

```{R, echo=FALSE, eval=FALSE}
# protein example (move this)

the protein exaple from @gauranEmpiricalNullEstimation2018
library(empiricalzip)
data(protein, package="empiricalzip")
k<-1
x <- protein[,(k+1)][!is.na(protein[,(k+1)])]
M <- length(x)
cf <- 10^(-10)
count <- c(length(x[which(x==0)]), tabulate(x, nbins = max(x)))
names(count) <- c(0:max(x))
aa <- my.RealDataFDR(x, 0.1) #
##     ZIGP ZIP   GP  P         ZIGP ZIP   GP  P        ZIGP  ZIP     GP  P
##      one stage procedure      two stage             Storey's FDR
## [1,]  191  212  212  212      201  212  223  212     200   211   211   211
## [2,]  143  205  212  212      143  205  229  212     162   204   211   211

## [[2]]
## #        eta      lambda    theta      pi    C0   AD   
## [1,] 0.3245463 1.9167862 0.1415783 0.4576160    6    7
## [2,] 0.2288849 1.0948534        NA 0.3985044    3    6
## [3,]        NA 0.8082199 0.0000000 0.3944174    3    4
## [4,]        NA 0.8082199        NA 0.3944174    3    6

## [[3]]
##           [,1]      [,2]      [,3]      [,4] [,5] [,6]
## [1,] 0.2251642 2.1452087 0.5741445 0.6141275    4   36
## [2,] 0.2759650 1.3856234        NA 0.4244176    2    6
## [3,]        NA 0.7994067 0.0000000 0.3929166    2    3
## [4,]        NA 0.7994067        NA 0.3929166    2    6

# ZIGP parameters
eta <- 0.3245463
lambda <- 1.9167862 
theta <- 0.1415783

# Generate the plots and analysis
results <- plot_zigp_comparison(
  observed_counts = count,
  eta = eta,
  lambda = lambda, 
  theta = theta,
  max_display = 30,
  y_min = -15  # Extended y-axis
)

results$log_scale_tail
abline(v=6)

plot_normalized_mixture(
   eta = eta,
   lambda =  lambda, 
   theta = theta,
   pi0 = 1,
   n_binom = 15,
   p_binom = 0.2,
   shift = 7
)

@@

# Save the improved plot
#ggsave("./script3/zigp_log_scale_extended.png", results$log_scale, width = 10, height = 6)
#ggsave("zigp_ratio_plot.png", results$ratio_plot, width = 10, height = 6)

@@

```{r zigplogscaleextended, eval=TRUE, echo=FALSE, fig.cap='empiricalzip fitted ZIPG',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./script3/zigp_log_scale_extended.png")

@@

\clearpage
\FloatBarrier
## Local FDR Calculation
The local FDR (false discovery rate) at a particular count value is calculated as:

@@

# hurdle model implementation

library(pscl)  # For hurdle model implementation

# Fit a hurdle model
hurdle_model <- hurdle(
  x ~ 1,               # Formula (just intercept in this simple case)
  data = data.frame(x = x_sample),
  dist = "negbin",     # Distribution for non-zero counts
  zero.dist = "binomial" # Distribution for zero vs non-zero process
)

summary(hurdle_model)

```{R, echo=FALSE, eval=FALSE}
# what is this? remove it
# Custom function for extreme zero-inflation
extreme_zero_model <- function(x, min_count = 3) {
  # Calculate basic statistics
  zeros <- sum(x == 0)
  n <- length(x)
  p_zero <- zeros/n

  # Estimate significance threshold
  # For extreme zero-inflation, we can use empirical approach
  unique_counts <- sort(unique(x))
  freq_table <- table(x)

  # Simple rule: counts are significant when they occur with 
  # frequency less than 0.1% of non-zero observations
  rarity_threshold <- 0.001 * (n - zeros)
  significant <- unique_counts[unique_counts >= min_count & 
                              freq_table[as.character(unique_counts)] < rarity_threshold]

  return(list(
    p_zero = p_zero,
    significant_counts = significant
  ))
}

@@

#png("./script3/ZI_Zipf_plot.png")
results$plots$"ZI_Zipf"
#dev.off()
#png("./script3/ZIGP_plot.png")
results$plots$"ZIGP"
#dev.off()
#png("./script3/ZINB_plot.png")
results$plots$"ZINB"
#dev.off()

@@

```{R, zipfplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Power Law (Zipf) Distribution", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/ZI_Zipf_plot.png")
```

```{R, zigpplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Generalized Poisson", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/ZIGP_plot.png")
```

```{R, zinbplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Negative Binomial", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/ZINB_plot.png")
```

@@

- A true power law would show a straight line. Deviations suggest a different distribution.
```{R, echo=FALSE, eval=FALSE}
# Plot log(frequency) vs log(count)
png("./script3/power_law.png")
plot(log(counts[counts > 0]), log(freqs[counts > 0]), 
     xlab="log(count)", ylab="log(frequency)")
abline(lm(log(freqs[counts > 0]) ~ log(counts[counts > 0])))
dev.off()

@@

```{R, powerlaw, echo=FALSE, eval=TRUE, fig.cap="Plot of log(frequency) vs log(count). A power law would show a straight line", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/power_law.png")
```

>>>>

---
title: "The Rheumatoid Arthritis data  \nSupplementary File for  \n`Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests`"
author: "Hawlader Al-Mamun,  Rob Dunne"
date: "Wednesday, September 25, 2024"
output: 
    bookdown::pdf_document2:
     toc: true
     toc_depth: 3
     fig_caption: yes
     citation_package: biblatex
bibliography: 
header-includes:
  \usepackage{dcolumn}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage{biblatex}
  \usepackage{caption}
  \usepackage{subcaption}
---

@@

```
# load data

We have

@@

```{r, echo=TRUE, eval=FALSE}
library(lmtest)
library(ranger)

numIte  <-  1
t <- numIte
ntree <- 1000

@@

dt$pheno <- as.factor(dt$pheno)
#dt <- dt[which(dt$treatment=="short"),-2]
```

```{r, echo=FALSE, eval=FALSE}
# quick load of results etc if we have a
dirOut  <-  paste0("Results")
load(file="Results/var_importance_gini.Rdata")
load(file=paste0(dirOut,"/All_Interactions.Rdata"))
load(file="./Results/interaction_snps_counts_fixed.Rdata")
load(file=paste0(dirOut,"/lm_results_fixed.Rdata"))

@@

```{r, echo=FALSE, eval=FALSE}
set.seed(100)
system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#89397.171     4.807  1591.941 
#save(model.rf, file="model.rf_temp3.Rdata")
var_importance <- model.rf$variable.importance
head(rev(sort(var_importance)))
t2 <-count_variables(model.rf)
names(t2) <-names(var_importance)

@@

```{r RFlocalfdr, eval=TRUE, echo=FALSE, fig.cap='RFlocalfdr analysis',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./supp_figs/RFlocalfdr.png")

@@

```{R, echo=TRUE, eval=FALSE}
library(RFlocalfdr)
vignette(package="RFlocalfdr")
imp<-log(var_importance)  #6275 var_importance vales =0 
plot(density(imp),xlab="log importances",main="")

@@

cutoffs <- c(4,5,6)
res.con2 <- determine_cutoff(imp,t2 ,cutoff=cutoffs,plot=c(4,5,6))
plot(c(0,1,2,3,4,5,6),c(res.con[,3],res.con2[,3]),pch=15,col="red",cex=1.5,
     ylab="max(abs(y - t1))")

@@

png("./supp_figs/RFlocalfdr.png")
aa<-significant.genes(ppp,temp,cutoff=0.05,debug.flag=0,do.plot=TRUE,use_95_q=TRUE)
dev.off()
length(aa$probabilities) # 74

@@

```{r hotspot, eval=TRUE, echo=FALSE, fig.cap='There is a hotspot in chromosome 6',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./supp_figs/hotspot.png")

@@

# count the interactions in the data

```{R, echo=TRUE, eval=FALSE}
for(i in 1:ntree){
    #tree2 <- getTree(model.rf2,i,T)
    tree <- treeInfo(model.rf,i)
    #tree <- treeInfo(model.n,5)

@@

#save(All_Interactions,file=paste0(dirOut,"/All_Interactions.Rdata"))
#dim(All_Interactions) #[1] 294310      2

```{R, echo=FALSE, eval=FALSE}
# just some checks on how  All_Interactions is organized. 
head(All_Interactions)
#            X1           X2      294310      2
#1  X6_32471505  X6_32771829
#2 X21_44075267  X6_32471505
#3 X14_84754343  X4_67191986

@@

@@

#save(interaction_snps_counts, file="./Results/interaction_snps_counts_fixed.Rdata")
#load(file="./Results/interaction_snps_counts_fixed.Rdata")

@@

## data set up

@@

## empiricalzip

@@

@@

# Save the improved plot
#ggsave("./supp_figs/zigp_log_scale_extended.png", results$log_scale, width = 10, height = 6)
#ggsave("zigp_ratio_plot.png", results$ratio_plot, width = 10, height = 6)

@@

```{r zigplogscaleextended, eval=TRUE, echo=FALSE, fig.cap='empiricalzip fitted ZIPG',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./supp_figs/zigp_log_scale_extended.png")

@@

## Local FDR Calculation
The local FDR (false discovery rate) at a particular count value is calculated as:

@@

@@

#png("./supp_figs/ZI_Zipf_plot.png")
results$plots$"ZI_Zipf"
#dev.off()
#png("./supp_figs/ZIGP_plot.png")
results$plots$"ZIGP"
#dev.off()
#png("./supp_figs/ZINB_plot.png")
results$plots$"ZINB"
#dev.off()

@@

```{R, zipfplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Power Law (Zipf) Distribution", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./supp_figs/ZI_Zipf_plot.png")
```

```{R, zigpplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Generalized Poisson", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./supp_figs/ZIGP_plot.png")
```

```{R, zinbplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Negative Binomial", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./supp_figs/ZINB_plot.png")
```

@@

- A true power law would show a straight line. Deviations suggest a different distribution.
```{R, echo=FALSE, eval=FALSE}
# Plot log(frequency) vs log(count)
png("./supp_figs/power_law.png")
plot(log(counts[counts > 0]), log(freqs[counts > 0]), 
     xlab="log(count)", ylab="log(frequency)")
abline(lm(log(freqs[counts > 0]) ~ log(counts[counts > 0])))
dev.off()

@@

```{R, powerlaw, echo=FALSE, eval=TRUE, fig.cap="Plot of log(frequency) vs log(count). A power law would show a straight line", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./supp_figs/power_law.png")
```
